---
title: "Assignment 5: Data Visualization"
author: "Cara Kuuskvere"
date: "Fall 2023"
output: pdf_document
geometry: margin=2.54cm
fig_height: 3
fig_width: 4
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1.  Rename this file `<FirstLast>_A05_DataVisualization.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.

---

## Set up your session 

1. Set up your session. Load the tidyverse, lubridate, here & cowplot packages, and verify your home directory. Read in the NTL-LTER processed data files for nutrients and chemistry/physics for Peter and Paul Lakes (use the tidy `NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv` version in the Processed_KEY folder) and the processed data file for the Niwot Ridge litter dataset (use the `NEON_NIWO_Litter_mass_trap_Processed.csv` version, again from the Processed_KEY folder). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r initialize project}
#1 
library(tidyverse)
library(cowplot)
library(lubridate)
library(ggthemes)
library(here)
here()
NEON_NIWO_Litter_mass_trap_Processed <- read.csv(here("Data/New Processed_key/NEON_NIWO_Litter_mass_trap_Processed.csv"), 
                          stringsAsFactors = FALSE)
NTL_PeterPaul <- read.csv(here("Data/New Processed_Key/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv"), 
                          stringsAsFactors = FALSE)
#2 
is.Date(NEON_NIWO_Litter_mass_trap_Processed$collectDate) #returns false
is.Date(NTL_PeterPaul$sampledate) #returns false
# Using lubridate ymd function
NEON_NIWO_Litter_mass_trap_Processed$collectDate <- 
  ymd(NEON_NIWO_Litter_mass_trap_Processed$collectDate)
NTL_PeterPaul$sampledate <-
  ymd(NTL_PeterPaul$sampledate)
# Check to verify it is now a date
is.Date(NTL_PeterPaul$sampledate)
is.Date(NEON_NIWO_Litter_mass_trap_Processed$collectDate)
```
## Define your theme
3. Build a theme and set it as your default theme. Customize the look of at least two of the following: 
 * Plot background
 * Plot title
 * Axis labels
 * Axis ticks/gridlines
 * Legend

```{r build plot theme}
#3 Building a custom theme for my legend and title
my_theme <- theme(legend.position = "top",
                  plot.title = element_text(size = 12),
                  legend.title = element_text(size = 12))
```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with separate aesthetics for Peter and Paul lakes. Add a line of best fit and color it black. Adjust your axes to hide extreme values (hint: change the limits using `xlim()` and/or `ylim()`).

```{r plot total P vs PO4}
#4  Plotting total phosphorus by phospate

 Total_tp_by_PO4 <- ggplot(NTL_PeterPaul,
      aes(
      x=po4,
      y=tp_ug,
      #separating aes for peter and paul lakes 
      color=lakename)
      )+
  geom_point()+
  my_theme+
  #Adding line of best fit
  geom_smooth(method=lm, formula = y~x, color='black')+
  #Adjusting x and y axes to hide extreme values 
  xlim(0,50)+
  ylim(0,100)+
  ggtitle("Total Phosphorus vs Phosphate in Peter and Paul Lakes")+
  labs(y="Phosphorus", x="Phosphate", color="Lake")
 
Total_tp_by_PO4
```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned. 

Tip:
* Recall the discussion on factors in the previous section as it may be helpful here. 
* R has a built-in variable called `month.abb` that returns a list of months;see <https://r-lang.com/month-abb-in-r-with-example>

```{r Create boxplots}
#5 
#Temperature box plot
BoxPlot_Temp <- ggplot(NTL_PeterPaul,
       aes(
         #Making x axis values the month abbreviations 
        x=factor(month, levels=1:12, labels=month.abb),
        y=temperature_C,
        color=lakename)
      )+
  #calling it as a box plot 
  geom_boxplot()+
  #Adding descriptive elements to the box plot
  ggtitle("Lake Temperature by Month")+
  labs(y="Temperature in Degrees C", x="Month")+
  #applying my theme
  my_theme

BoxPlot_Temp

#TP Box plot 
BoxPlot_TP <- ggplot(NTL_PeterPaul,
       aes(
        #Making x axis values the month abbreviations 
        x=factor(month, levels=1:12, labels=month.abb,
                 ordered = is.ordered(month.name)),
        y=tp_ug,
        color=lakename)
      )+
  #calling it as a box plot
  geom_boxplot()+
  #adding descriptive elements
  ggtitle("Lake Total Phosophorus by Month")+
  labs(y="Total Phosophorus", x="Month")+
  #applying my theme
  my_theme

BoxPlot_TP

#TN Box plot
BoxPlot_TN <- ggplot(NTL_PeterPaul,
       aes(
        #Making x axis values the month abbreviations     
        x=factor(month, levels=1:12, labels=month.abb),
        y=tn_ug,
        color=lakename)
      )+
  #calling it as a box plot
  geom_boxplot()+
  #adding descriptive information to the chart
  ggtitle("Lake Total Nitrogen by Month")+
  labs(y="Total Nitrogen", x="Month")+
  #applying my custom theme 
  my_theme

BoxPlot_TN

#Creating a cowplot to combine the three graphs 

Combined_Temp_TP_TN <- plot_grid(BoxPlot_Temp,BoxPlot_TN,BoxPlot_TP, 
                                 ncol=1, nrow = 3, align=c("v"),
                                 axis=factor(month, levels=1:12, 
                                             labels=month.abb)
                                  
                                 )
                               
                
      
Combined_Temp_TP_TN  


```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: Paul lake has more measurements of these variables of interested in the early summer/ late spring, and Peter lake has more measurements and prevelence in the late summer/ early fall. The IQR of nitrogen and phosophorus is much smaller than the IRQ temperature. Futher, the temperature of the lakes fall a small amount from mid summer to mid fall. 

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r Plot litter}
#6
Litter_Plot <- NEON_NIWO_Litter_mass_trap_Processed %>% 
  filter(functionalGroup == "Needles") %>% 
  ggplot(
    mapping = aes(
      x=collectDate,
      y=dryMass,
      color=nlcdClass)
    ) + 
  geom_point()+
  ggtitle("Dry Mass of Needle Litter by Date and NLCD Class")+
  labs(y="Dry Mass", x="Year")
Litter_Plot

#7
Litter_Plot_Sep <- NEON_NIWO_Litter_mass_trap_Processed %>% 
  filter(functionalGroup == "Needles") %>% 
  ggplot(
    mapping = aes(
      x=collectDate,
      y=dryMass)
    ) + 
  facet_wrap(vars(nlcdClass))+
  geom_point()+
  ggtitle("Dry Mass of Needle Litter by Date")+
  labs(y="Dry Mass", x="Year")
Litter_Plot_Sep

```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer: Plot 7 is more effective becasue you are able to see the individual relationships between points of different nlcd locations, and these differences are more clear. In plot 6, it is difficult to see the data trends because you have to focus much harder on sepearating out the locations and colors you are looking at. I think if you want an overall look at dry mass versus collection date, six would be more effective and offers more information on the nlcd class, but for a purpose of seeing the differences between the classes, #7 plot is best. 