---
title: "Assignment 2: Coding Basics"
author: "Cara Kuuskvere"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Sakai.

## Basics, Part 1

1.  Generate a sequence of numbers from one to 30, increasing by threes. Assign this sequence a name.

2.  Compute the mean and median of this sequence.

3.  Ask R to determine whether the mean is greater than the median.

4.  Insert comments in your code to describe what you are doing.

```{r basics 1}
#1. I created an object named "basics1" that is a sequence of numbers from 1 to 30 increasing by 3's. I then printed this sequence (and commented this action out) to review its outputs and check my work. 
basics1 <- seq (1,30,3)
#print(basics1)

#2. I created two objects, one named "basicsmean" which is assigned the mean value of the sequence "basics1". I then created a "basicsmedian" object, which is assigned the value fo the median of the sequence "basics1." I printed each of these to see their values, which I commented out since that was not requested. 
basicsmean <- mean(basics1)
print (basicsmean)
basicsmedian <- median(basics1)
#print (basicsmedian)
#3. I ran a boolean test to see if "basicsmean" is greater than "basicsmedian", and this returned the value of false. I also ran the boolean to see if the mean is less than the median, which also returned false -- therefore, the values are equal. 
basicsmean > basicsmedian
basicsmean < basicsmedian
```

## Basics, Part 2

5.  Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6.  Label each vector with a comment on what type of vector it is.

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

8.  Label the columns of your data frame with informative titles.

```{r basics 2}
student_name <- c("hannah", "nicole", "zach", "cara")
test_scores <- c(65, 95, 30, 77)
passing_score <- c(test_scores>=50)


test_grader <- data.frame(student_name, test_scores, passing_score)
colnames(test_grader) <- c("Student Name", "Exam Score out of 100", "Passing?")

#print(test_grader)


```

9.  QUESTION: How is this data frame different from a matrix?

> Answer: A data frame and a matrix both consist of a list of vectors of equal length. However, a in a matrix all the vectors must be of the same type (ie character, integer, etc.) whereas in a data frame the vectors do not need to be the same type (you can have a vector of characters accompanied by a vector of integers). 

10. Create a function with an if/else statement. Your function should take a **vector** of test scores and print (not return) whether a given test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. 

11. Apply your function to the vector with test scores that you created in number 5.

```{r basics 3}
#grader2 <- function(x){
# if (x>=50){
#    print("TRUE")
#  }
#  else {
#    print("FALSE")
#  }
#}

#grader2(3)
#grader2(test_scores)

grader <- function (x) {ifelse(x>=50,"PASSING","FAILING")
}
print(grader(test_scores))
```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: "ifelse" worked rather than "if, else." Ifelse returned the output of the grader function because it is able to work with vectors. However, I received an error for my condition of the if, else statement that it has length greater than 1, due to inputting a vector for "X". I tested my if else with an integer, and the function worked. My if, else statement, which is commented out, can be un-commented and run to demonstrate this.