---
title: "Assignment 4: Data Wrangling"
author: "Cara Kuuskvere"
output: pdf_document
geometry: margin=2.54cm
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Wrangling 

## Directions

1. Rename this file `<FirstLast>_A04_DataWrangling.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.
6. Ensure that code in code chunks does not extend off the page in the PDF.

The completed exercise is due on Thursday, Sept 28th @ 5:00pm.

## Set up your session 

1a. Load the `tidyverse`,  `lubridate`, and `here` packages into your session.

1b. Check your working directory.

1c. Read in all four raw data files associated with the EPA Air dataset, being sure to set string columns to be read in a factors. See the README file for the EPA air datasets for more information (especially if you have not worked with air quality data previously). 

2. Apply the `glimpse()` function to reveal the dimensions, column names, and structure of each dataset. 

```{r setup packages, wd, and files, message = FALSE}
#1a loaded necessary packages
#no need to run install as I already have these loaded, but can be uncommented 
#install(tidyverse)
#install(lubridate)
#install(here)
#install(dplyr)
library(tidyverse)
library(lubridate)
library(here)
library(dplyr)
#1b checking my working directoy 
here()
# can also use getwd()

#1c loading each dataset using read_csv
EPAair_O3_NC2018_raw <- read_csv(file=here('Data/Raw/EPAair_O3_NC2018_raw.csv'),
                                  (stringsAsFactors = TRUE))
EPAair_O3_NC2019_raw <- read_csv(here('Data/Raw/EPAair_O3_NC2019_raw.csv'),
                                  (stringsAsFactors = TRUE))
EPAair_PM25_NC2018_raw <- read_csv(here('Data/Raw/EPAair_PM25_NC2018_raw.csv'),
                                  (stringsAsFactors = TRUE))
EPAair_PM25_NC2019_raw <- read_csv(here('Data/Raw/EPAair_PM25_NC2019_raw.csv'),
                                  (stringsAsFactors = TRUE))

EPAair_O3_NC2018_raw
EPAair_O3_NC2019_raw
EPAair_PM25_NC2018_raw
EPAair_PM25_NC2019_raw
#2 using glimpse for each dataset
glimpse(EPAair_O3_NC2018_raw)
glimpse(EPAair_O3_NC2019_raw)
glimpse(EPAair_PM25_NC2018_raw)
glimpse(EPAair_PM25_NC2019_raw)
```


## Wrangle individual datasets to create processed files.

3. Change the Date columns to be date objects.

4. Select the following columns: Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, COUNTY, SITE_LATITUDE, SITE_LONGITUDE

5. For the PM2.5 datasets, fill all cells in AQS_PARAMETER_DESC with "PM2.5" (all cells in this column should be identical).

6. Save all four processed datasets in the Processed folder. Use the same file names as the raw files but replace "raw" with "processed".

```{r Wranlging individual datasets to create processed files}
#3 Using the mdy function from lubridate to change the data columns from
#characters to date objects

EPAair_O3_NC2018_raw$Date <- mdy(EPAair_O3_NC2018_raw$Date)
EPAair_O3_NC2019_raw$Date <- mdy(EPAair_O3_NC2019_raw$Date)
EPAair_PM25_NC2018_raw$Date <- mdy(EPAair_PM25_NC2018_raw$Date)
EPAair_PM25_NC2019_raw$Date <- mdy(EPAair_PM25_NC2019_raw$Date)

#4 using a pipe and "select" function 
EPAair_O3_NC2018_processed <- EPAair_O3_NC2018_raw %>% 
  select(Date, DAILY_AQI_VALUE, "Site Name", AQS_PARAMETER_DESC, COUNTY, 
         SITE_LATITUDE, SITE_LONGITUDE)
EPAair_O3_NC2019_processed <- EPAair_O3_NC2019_raw %>% 
  select(Date, DAILY_AQI_VALUE, "Site Name", AQS_PARAMETER_DESC, COUNTY, 
         SITE_LATITUDE, SITE_LONGITUDE)
EPAair_PM25_NC2018_processed <- EPAair_PM25_NC2018_raw %>% 
  select(Date, DAILY_AQI_VALUE, "Site Name", AQS_PARAMETER_DESC, COUNTY, 
         SITE_LATITUDE, SITE_LONGITUDE) %>% 
  #Q5
  mutate(AQS_PARAMETER_DESC = "PM 2.5")
EPAair_PM25_NC2019_processed <- EPAair_PM25_NC2019_raw %>% 
  select(Date, DAILY_AQI_VALUE, "Site Name", AQS_PARAMETER_DESC, COUNTY, 
         SITE_LATITUDE, SITE_LONGITUDE) %>% 
  #Q5
  mutate(AQS_PARAMETER_DESC = "PM 2.5")

#5 see pipe above for "mutate" function on the PM2.5 datasets

#calling each processed individual dataset
EPAair_O3_NC2018_processed
EPAair_O3_NC2019_processed
EPAair_PM25_NC2018_processed
EPAair_PM25_NC2019_processed

#6 Using write.csv, not inlcuding row names, and saving to the file path in my
#WD where the processed data is
write.csv(EPAair_O3_NC2018_processed, row.names= FALSE,
        file="./Data/Processed/EPAair_O3_NC2018_processed.csv")
write.csv(EPAair_O3_NC2019_processed, row.names = FALSE,
        file = "./Data/Processed/EPAair_O3_NC2019_processed.csv")
write.csv(EPAair_PM25_NC2018_processed, row.names =FALSE, 
        file ="./Data/Processed/EPAair_PM25_NC2018_processed.csv")
write.csv(EPAair_PM25_NC2019_processed, row.names=FALSE,
        file= "./Data/Processed/EPAair_PM25_NC2019_processed.csv")


```


## Combine datasets

7. Combine the four datasets with `rbind`. Make sure your column names are identical prior to running this code.

8. Wrangle your new dataset with a pipe function (%>%) so that it fills the following conditions: 

  * Include only sites that the four data frames have in common: "Linville Falls", "Durham Armory", "Leggett", "Hattie Avenue", "Clemmons Middle", "Mendenhall School", "Frying Pan Mountain", "West Johnston Co.", "Garinger High School", "Castle Hayne", "Pitt Agri. Center", "Bryson City", "Millbrook School" (the function `intersect` can figure out common factor levels - but it will include sites with missing site information, which you don't want...) 

  * Some sites have multiple measurements per day. Use the split-apply-combine strategy to generate daily means: group by date, site name, AQS parameter, and county. Take the mean of the AQI value, latitude, and longitude.
  
  * Add columns for "Month" and "Year" by parsing your "Date" column (hint: `lubridate` package)
  
  * Hint: the dimensions of this dataset should be 14,752 x 9.

9. Spread your datasets such that AQI values for ozone and PM2.5 are in separate columns. Each location on a specific date should now occupy only one row.

10. Call up the dimensions of your new tidy dataset.

11. Save your processed dataset with the following file name: "EPAair_O3_PM25_NC1819_Processed.csv"

```{r Combining datasets and filtering/cleaning combined data}
#7 Using rbind to combine the four datasets. Verifed column names are identical. 

EPAair_O3_PM25_NC_1819_raw <- rbind(EPAair_O3_NC2018_processed, EPAair_O3_NC2019_processed, EPAair_PM25_NC2018_processed, EPAair_PM25_NC2019_processed)

# calling my new combined but raw dataset
EPAair_O3_PM25_NC_1819_raw

#8 Processing using a pipe

EPAair_O3_PM25_NC_1819_processed <- 
  # specify my dataframe is the raw combined data
  EPAair_O3_PM25_NC_1819_raw %>% 
  # filtering for the sites that are in common using "or" operator 
  filter(`Site Name` == "Linville Falls" |`Site Name`== "Durham Armory" | `Site Name`== "Leggett"| `Site Name`== "Hattie Avenue"| `Site Name`== "Clemmons Middle"|`Site Name`== "Mendenhall School"| `Site Name`== "Frying Pan Mountain"| `Site Name`== "West Johnston Co."| `Site Name`== "Garinger High School"| `Site Name`== "Castle Hayne"| `Site Name`== "Pitt Agri. Center"| `Site Name`== "Bryson City"| `Site Name`== "Millbrook School") %>% 
  # removing na values so I can calculate the mans 
  filter(!is.na(DAILY_AQI_VALUE) & !is.na(SITE_LATITUDE) & !is.na(SITE_LONGITUDE)) %>%
  # grouping by the specified groups 
  group_by(Date,`Site Name`, COUNTY, AQS_PARAMETER_DESC) %>%
  #using summarise to find the means to AQI, lat, and long 
  summarise(meanAQI = mean(DAILY_AQI_VALUE), 
            meanLAT = mean(SITE_LATITUDE), 
            meanLONG = mean(SITE_LONGITUDE)) %>%  
  # mutating to add a month and year cloumn
  mutate(month = month(Date)) %>% 
  mutate(year= year(Date)) %>% #at this point, before the pipe and spread the dimensions at 14752 by 9
#9 final part of pipe, spreading to create two cloumns for ozone and PM 2.5
  spread(AQS_PARAMETER_DESC, meanAQI)

#calling this new df
EPAair_O3_PM25_NC_1819_processed

#10 calling dimensions of my processed data

dim(EPAair_O3_PM25_NC_1819_processed)

#11 saving as a CSV in my processed data folder using the file path 

write.csv(EPAair_O3_PM25_NC_1819_processed,row.names= FALSE,
          file = "./Data/Processed/EPAair_O3_PM25_NC_1819_processed.csv")
```


## Generate summary tables

12. Use the split-apply-combine strategy to generate a summary data frame. Data should be grouped by site, month, and year. Generate the mean AQI values for ozone and PM2.5 for each group. Then, add a pipe to remove instances where mean **ozone** values are not available (use the function `drop_na` in your pipe). It's ok to have missing mean PM2.5 values in this result. 

13. Call up the dimensions of the summary dataset.

```{r Generating Summary Tables of O3 PM2.5 NC 1819 Processed DF }

#12 Using split-combine-apply to summarize the data frame grouped by site name,
# year, and month, removing NA values, and summarizing using the mean fn

EPAair_O3_PM25_NC_1819_processed_summaries <- 
  EPAair_O3_PM25_NC_1819_processed %>% 
  group_by(`Site Name`,year, month, .add=FALSE) %>% 
  drop_na(Ozone) %>%
  summarise(meanAQI_Ozone =mean(Ozone),
            meanAQI_PM2.5 = mean(`PM 2.5`))
#calling my summary df
EPAair_O3_PM25_NC_1819_processed_summaries

#13 Calling the dimensions of my summarized processed df
dim(EPAair_O3_PM25_NC_1819_processed_summaries)

```

14. Why did we use the function `drop_na` rather than `na.omit`?

> Answer: na.omit will remove all rows that have at least one na, and we are able to keep the PM2.5 rows that have na data. Drop_na allows us to drop only the rows that have an na in the specific ozone column, whereas na.omit will drop more rows than we want dropped. 
